Easy as Pi, Volume II: On the Goodput and Energy of Network Coding Schemes in the Raspberry Pi

### 2016-08-10 schemes.tex
Changing introduction...
minor corrections

In this section, we present a description of the considered coding
schemes measured with the \ac{Raspi} testbed. We introduce a
definition for the primitive coding operations, e.g. encoding,
decoding and recoding (where it applies) for each coding
scheme. Later, we address particular schemes which are obtained by
modifying the basic coding operations that provide better processing
speeds, which is particularly relevant for the \ac{Raspi}.  Finally,
we include a review of algorithms for network coding that exploit
multicore capabilities of the \ac{Raspi} 2.

@@

\subsubsection{Encoding}
In \ac{RLNC}, any coded packet is a linear combination from all
the original packets. For the coding scheme, packets are seen as
algebraic entities formed as a sequence of elements from $GF(q)$,
which is a \ac{GF} of size $q$. Later, each original packet is
multiplied by a coding coefficient from $GF(q)$. The coding coefficients
are chosen uniformly at random from the \ac{GF} by the encoder. To
perform the multiplication of a packet by a coding coefficient, the
coefficient is multiplied for each of the elements in the
concatenation that composes an original packet, preserving the
concatenation. Later, all resulting packets are added within the
\ac{GF} arithmetics together to generate a coded packet. Thus, a
coded packet can be written as:
%
\begin{align} \label{eq:coded_packet}
C_i  &= \bigoplus_{j=1}^{g} v_{ij} \otimes P_j ,\ \forall i \in [1,\ldots)
\end{align}

@@

A decoder starts to calculate and substract contributions from each
of the pivot elements, e.g. leftmost elements in the main diagonal of
\eqref{eq:coding_matrix}, from top to bottom. The purpose is to obtain
the equivalence  $\textbf{V}$ in its reduced echelon form. The steps
for reducing the matrix by elementary row operations are carried each time
a \ac{l.i.} packet is received. Once in reduced echelon form, packets
can be retrieved by doing a substitution starting from the latest
coded packet. In this way, the amount of elementary operations at the end
of the decoding process is diminished.

\subsubsection{Recoding}
As an inherent property of \ac{RLNC}, an intermediate node in the network
is able to create new coded packets without needing to decode previously
received ones from an encoding source. Therefore, \ac{RLNC} is an end-to-end
coding scheme that permits to \textit{recode} former coded packets at any
point in the network without requiring a local decoding of the data. In
principle, a recoded packet should be indistinguishable from a coded one.
Thus, we define a recoded packet as $R_i$ and consider the coding
coefficients $w_{i1}, \ldots, w_{ig}$ as the used ones to
create $R_i$. Later, a recoded packet can be written as:
%
\begin{align}
\label{eq:recoded_packet}
R_i = \bigoplus_{j=1}^{g} w_{ij} \otimes C_j  ,\ \forall i \in [1,\ldots).
\end{align}

@@

In \ac{SRLNC}, instead of considering all the packets to create a
coded packet as in \ac{RLNC}, an encoder sets more coding coefficients
to zero when generating a coded packet with the purpose of reducing
the overall processing.  Decoding is the same as in \ac{RLNC}, but
given that the coding matrices are now sparse, we will have less
operations performed in the decoding process.  Recoding, although
theoretically possible, is omitted since it requires the use of
heuristics keep packets sparse after recoding, which is inherently
sub-optimal. In what follows, we describe the coding scheme with a
basic method to produce sparse coded packets and later we indicate a
second one, giving reference to their implementation performance.

\subsubsection{Method 1: Fixing the Coding Density}
A way to control the amount of non-zero coding coefficients, is to set a
fixed ratio of non-zero coefficients in the encoding vector of size $g$. We
refer to this fixed ratio as the average coding density $d$. Thus, for any
coded packet $C_i$ with coding coefficients $v_{ij},\ j \in [1,g]$, its
coding density is defined as follows:
%
\begin{align}
\label{eq:coding_density}
d = \frac{\sum_{j=1}^{g} f(v_{ij})}{g},\ \
    f(v_{ij}) =
    \begin{cases}
        0 &,\ v_{ij} = 0 \\
        1 &,\ v_{ij} \neq 0
    \end{cases}
\end{align}
%
From the density definition in~\eqref{eq:coding_density}, it can be
observed that $0 \leq d \leq 1$. As $g$ increased, we obtain more
granularity in the density. Notice that the special case of  $d = 0$ has no
practical purpose since it implies just zero padded data. So, a practical
range for the density excludes the zero case, $0 < d \leq 1$. Moreover,
it can be also observed that the case of $d = 1$ reduces to the \ac{RLNC}
coding scheme described in Section~\ref{ssec:RLNC}.

To achieve a desired average coding density in the coefficients, for each
of them, we utilize a set of Bernoulli random variables all with parameter
$d$ as it success probability, i.e. $\mathbb{B}_j \sim Bernoulli(d)\ ,\ \forall
j \in [1,g]$. In this way, we can represent a coded packet in \ac{SRLNC} as:
%
\begin{align} \label{eq:sparse_coded_packet}
    C_i  &= \bigoplus_{j=1}^{g} \mathbb{B}_j v_{ij} \otimes P_j ,\ v_{ij}
    \neq 0\
    \forall i \in [1,\ldots) ,\ \
    d \in
    \begin{cases}
        {(0,0.5]} & ,\ q = 2 \\
        {(0,1]}   & ,\ q > 2
    \end{cases}
\end{align}
%
In~\eqref{eq:sparse_coded_packet}, we have the requirement for the
coding coefficient to not be zero, since we want to ensure a coding
coefficient is generated for any random trial, where
$\mathbb{B}_j = 1$. Therefore, in our implementation of \ac{SRLNC} we
exclude the zero element and then pick uniformly distributed random
elements from $GF(q)-\{ 0 \}$. Also, we have specified a dependency on
the field size for practical density values. In case of employing
$GF(2)$, the maximum plausible density is restricted up to 0.5 since
higher values incur in more frequent \ac{l.d.} coded
packets~\cite{realworld_nc2013}.

Reducing $d$ enables the encoder to decrease the average number of
packets manipulated to make a coded one. This reduces the complexity
of the encoder since it needs to mix less packets. Moreover, it also
simplifies the decoder processing given that less nonzero coding
coefficients are required to be operated during the Gaussian
elimination stage.

@@

\subsubsection{Method 2: Sampling the Amount of Packets to Combine}
The method described from \eqref{eq:sparse_coded_packet} results
in a fast implementation in terms of execution time for
$d \geq 0.3$~\cite{practicalview_tsnc2015}. Nevertheless, for $d < 0.3$
the implementation becomes slow given that more Bernoulli trials are
required in total for all random variables to generate a coded packet.
Thus, we introduce a second method that permits a fast implementation
for low coding densities~\cite{practicalview_tsnc2015}.

@@

The only caveat is that the case of $\mathbb{K} = 1$ occurs slightly
more often than $\mathbb{N} = 1$ in the original distribution, but for
the considered density range in this method, this is not a significant
modification~\cite{practicalview_tsnc2015}. Then, once the
distribution for the number of packets to mix has been defined, we
sample a value $n$ from $\mathbb{N}$ and compute $k =
\max(1,n)$. Later, we create a create a set $\mathcal{K}$ with
cardinality $k$, e.g. $|\mathcal{K}| = k$, where the elements in
$\mathcal{K}$ are the indexes of the packets that are going to be
considered for making a coded packet. To compute the indexes
of the set $\mathcal{K}$, we do the following algorithm in pseudo-code: \\
\ \\
\begin{algorithm}[H]
 \label{alg:k_set}
 \KwData{$k$: Size of $\mathcal{K}$. $g$: Generation Size}
 \KwResult{$\mathcal{K}$: The set of non-repeated indexes}
 $\mathcal{K}$ = \{ \}\;
 \While{$|\mathcal{K}| \neq k$ }{
  $i$ = Sample from $\mathbb{U}(1,g)$\;
  \If{$i \notin \mathcal{K}$}{
   Insert $i$ in $\mathcal{K}$\;
   }
 }
 \caption{Computation of the set of indexes for packet combination in SRLNC.}
\end{algorithm}
\ \\
In Algorithm~\ref{alg:k_set}, the notation $\mathbb{U}(1,g)$ stands
for a uniform discrete random variable with limits $1$ and $g$. The
pseudo-code in Algorithm~\ref{alg:k_set} indicates that the set $\mathbb{K}$
is filled with non-repeated elements taken uniformly at random from
$[1,\ldots,g]$. Finally, once the set of indexes has been defined, a coded
packet using \ac{SRLNC} can be written as shown in
\eqref{eq:sparse_coded_packet2}:
%
\begin{align} \label{eq:sparse_coded_packet2}
    C_i  &= \bigoplus_{m \in \mathcal{K}} v_{im} \otimes P_{m},\ v_{im} \neq 0,\ \forall i \in [1,\ldots)
    %L \subseteq \{1,\dots,g\}: |L|=k ,\
\end{align}

% It is often desired to express the coding density as $d=\frac{k}{g}$. We call
% this notation $k-sparse$ where $k$ is the average number of packets combined to
% produce a coded packet.

%For low coding density, the encoder might generate a coded packet that consists
%of no coded packets. I.e. the packet consists of all zeroes

%In \ac{SRLNC}, instead of considering all the packets to create a coded
%packet like \ac{RLNC} does, here the encoder sets some coding coefficients to
%zero when creating a new coded packet. This permits to reduce the complexity
%of both the encoding and decoding operations since the encoder (or decoder)
%requires less operations to process the data. Therefore, a new code parameter
%to control the sparsity, the code density $d$, is included to vary the
%number of non-zero packet that are combined. Here are various approaches to
%set the density.
%
%A first method is to specify the number of zeros in a given
%generation size and later generate coded packets with this amount. For example:
%16 zeros in a coding vector of 64 packets where the position of zeros are
%modified at random for every coded packet generated. Another method is to
%define an average number of zeros in the coding vector. Therefore, more or
%less zeros are exactly generated for each coded packet in the process.
%However, if not carefully made, using a sparse coding scheme might increase
%the probability of generating \ac{l.d.} packets which incurs both overhead
%from redundant information and processing delay.

>>>>

In this section, we present the considered coding schemes that are evaluated
in the \ac{Raspi} 1 and 2.
%In this section, we present a description of the considered coding
%schemes measured with the \ac{Raspi} testbed.
We introduce a
definition for the primitive coding operations, e.g. encoding,
decoding and recoding (where it applies) for each coding
scheme. Later, we address particular schemes which are obtained by
modifying the basic coding operations that provide better processing
speeds, which is particularly relevant for the \ac{Raspi}.  Finally,
we include a review of algorithms for network coding that exploit
multicore capabilities of the \ac{Raspi} 2.

@@

\subsubsection{Encoding}
In \ac{RLNC}, any coded packet is a linear combination of all
the original packets. For the coding scheme, packets are seen as
algebraic entities formed as a sequence of elements from $GF(q)$,
which is a \ac{GF} of size $q$. Later, each original packet is
multiplied by a coding coefficient from $GF(q)$. The coding coefficients
are chosen uniformly at random from the \ac{GF} by the encoder. To
perform the multiplication of a packet by a coding coefficient, the
coefficient is multiplied for each of the elements in the
concatenation that composes an original packet, preserving the
concatenation. Later, all resulting packets are added within the
\ac{GF} arithmetics together to generate a coded packet. Thus, a
coded packet can be written as:
%
\begin{align} \label{eq:coded_packet}
C_i  &= \bigoplus_{j=1}^{g} v_{ij} \otimes P_j ,\ \forall i \in [1,\ldots)
\end{align}

@@

A decoder starts to calculate and substract contributions from each
of the pivot elements, e.g. leftmost elements in the main diagonal of
\eqref{eq:coding_matrix}, from top to bottom. The purpose is to obtain
the equivalence  $\textbf{V}$ in its reduced echelon form. The steps
for reducing the matrix by elementary row operations are carried out each time
a \ac{l.i.} packet is received. Once in reduced echelon form, packets
can be retrieved by doing a substitution starting from the latest
coded packet. In this way, the amount of elementary operations at the end
of the decoding process is diminished.

\subsubsection{Recoding}
As an inherent property of \ac{RLNC}, an intermediate node in the network
is able to create new coded packets without needing to decode previously
received packets from an encoding source. Therefore, \ac{RLNC} is an end-to-end
coding scheme that permits to \textit{recode} former coded packets at any
point in the network without requiring a local decoding of the data. In
principle, a recoded packet should be indistinguishable from a coded one.
Thus, we define a recoded packet as $R_i$ and consider the coding
coefficients $w_{i1}, \ldots, w_{ig}$ as used to
create $R_i$. Later, a recoded packet can be written as:
%
\begin{align}
\label{eq:recoded_packet}
R_i = \bigoplus_{j=1}^{g} w_{ij} \otimes C_j  ,\ \forall i \in [1,\ldots).
\end{align}

@@

In \ac{SRLNC}, instead of considering all the packets to create a
coded packet as in \ac{RLNC}, an encoder sets more coding coefficients
to zero when generating a coded packet with the purpose of reducing
the overall processing.  Decoding is the same as in \ac{RLNC}, but
given that the coding matrices are now sparse means that there will be less
operations to perform in the decoding process.  Recoding, although
theoretically possible, is omitted since it requires the use of
heuristics to keep packets sparse after recoding, which is inherently
sub-optimal.
%In what follows, we describe the coding scheme with a
%basic method to produce sparse coded packets and later we indicate a
%second one, giving reference to their implementation performance.
In what follows, we describe the coding scheme with two
different methods to produce sparse coded packets.

\subsubsection{Method 1: Fixing the Coding Density}
A way to control the amount of non-zero coding coefficients, is to set a
fixed ratio of non-zero coefficients in the encoding vector of size $g$. We
refer to this fixed ratio as the average coding density $d$. Thus, for any
coded packet $C_i$ with coding coefficients $v_{ij},\ j \in [1,g]$, its
coding density is defined as follows:
%
\begin{align}
\label{eq:coding_density}
d = \frac{\sum_{j=1}^{g} f(v_{ij})}{g},\ \
    f(v_{ij}) =
    \begin{cases}
        0 &,\ v_{ij} = 0 \\
        1 &,\ v_{ij} \neq 0
    \end{cases}
\end{align}
%
From the density definition in~\eqref{eq:coding_density}, it can be
observed that $0 \leq d \leq 1$. As $g$ increased, we obtain more
granularity in the density. Notice that the special case of  $d = 0$ has no
practical purpose since it implies just zero padded data.
\textcolor{red}{exclude the case d=1 for gf(2) here}
So, a practical
range for the density excludes the zero case, $0 < d \leq 1$.
%Moreover,
%it can be also observed that the case of $d = 1$ reduces to the \ac{RLNC}
%coding scheme described in Section~\ref{ssec:RLNC}.
Moreover, the densities $d=0.5$ in $GF(2)$ and $d=1$ in higher order
\ac{GF}s is similar to the \ac{RLNC} scheme described in Section~\ref{ssec:RLNC}.

To achieve a desired average coding density in the coefficients, for each
of them, we utilize a set of Bernoulli random variables all with parameter
$d$ as its success probability, i.e. $\mathbb{B}_j \sim Bernoulli(d)\ ,\ \forall
j \in [1,g]$. In this way, we can represent a coded packet in \ac{SRLNC} as:
%
\begin{align} \label{eq:sparse_coded_packet}
    C_i  &= \bigoplus_{j=1}^{g} \mathbb{B}_j v_{ij} \otimes P_j ,\ v_{ij}
    \neq 0\
    \forall i \in [1,\ldots) ,\ \
    d \in
    \begin{cases}
        {(0,0.5]} & ,\ q = 2 \\
        {(0,1]}   & ,\ q > 2
    \end{cases}
\end{align}
%
In~\eqref{eq:sparse_coded_packet}, we have the requirement for the
coding coefficient to not be zero, since we want to ensure a coding
coefficient is generated for any random trial, where
$\mathbb{B}_j = 1$. Therefore, in our implementation of \ac{SRLNC} we
exclude the zero element and then pick uniformly distributed random
elements from $GF(q)-\{ 0 \}$. Also, we have specified a dependency on
the field size for practical density values. In case of employing
$GF(2)$, the maximum plausible density is restricted up to 0.5 since
higher values incur in more frequent \ac{l.d.} coded
packets~\cite{realworld_nc2013}.\textcolor{red}{Rather more coding complexity}

Reducing $d$ enables the encoder to decrease the average number of
packets mixed to make a coded one. This reduces the complexity
of the encoder since it needs to mix less packets. Moreover, it also
simplifies the decoder processing given that less nonzero coding
coefficients are required to be operated during the Gaussian
elimination stage.

@@

\subsubsection{Method 2: Sampling the Amount of Packets to Combine}
The method described from \eqref{eq:sparse_coded_packet} results
in a fast implementation in terms of execution time for
$d \geq 0.3$~\cite{practicalview_tsnc2015}.
It is however not able to utilize the full performance potential for
low densities as the total number of Bernoulli trials remain unchanged
independently of the density.
%Nevertheless, for $d < 0.3$
%the implementation becomes slow given that more Bernoulli trials are
%required in total for all random variables to generate a coded packet.
Thus, we introduce a second method that permits a faster implementation
for low coding densities~\cite{practicalview_tsnc2015}.

@@

The only caveat is that the case of $\mathbb{K} = 1$ occurs slightly
more often than $\mathbb{N} = 1$ in the original distribution, but for
the considered density range in this method, this is not a significant
modification~\cite{practicalview_tsnc2015}. Then, once the
distribution for the number of packets to mix has been defined, we
sample a value $n$ from $\mathbb{N}$ and compute $k =
\max(1,n)$. Later, we create a create a set $\mathcal{K}$ with
cardinality $k$, e.g. $|\mathcal{K}| = k$, where the elements in
$\mathcal{K}$ are the indexes of the packets that are going to be
considered for making a coded packet. To compute the indexes
of the set $\mathcal{K}$, we do the following algorithm in pseudo-code: \\
\ \\
\begin{algorithm}[H]
 \label{alg:k_set}
 \KwData{$k$: Size of $\mathcal{K}$. $g$: Generation Size}
 \KwResult{$\mathcal{K}$: The set of non-repeated indexes}
 $\mathcal{K}$ = \{ \}\;
 \While{$|\mathcal{K}| \neq k$ }{
  $i$ = Sample from $\mathbb{U}(1,g)$\;
  \If{$i \notin \mathcal{K}$}{
   Insert $i$ in $\mathcal{K}$\;
   }
 }
 \caption{Computation of the set of indexes for packet combination in SRLNC.}
\end{algorithm}
\ \\
In Algorithm~\ref{alg:k_set}, the notation $\mathbb{U}(1,g)$ stands
for a uniform discrete random variable with limits $1$ and $g$. The
pseudo-code in Algorithm~\ref{alg:k_set} indicates that the set $\mathbb{K}$
is filled with non-repeated elements taken uniformly at random from
$[1,\ldots,g]$. Finally, once the set of indexes has been defined, a coded
packet using \ac{SRLNC} can be written as shown in
\eqref{eq:sparse_coded_packet2}:
%
\begin{align} \label{eq:sparse_coded_packet2}
    C_i  &= \bigoplus_{m \in \mathcal{K}} v_{im} \otimes P_{m},\ v_{im} \neq 0,\ \forall i \in [1,\ldots)
\end{align}
